---
title: "PM566 Final Project - Leading Causes of Death: United States"
author: "Jiayi Nie"
date: 12/03/2021
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

Last updated: `r format(Sys.time(), '%d %B, %Y')`

<br>
```{r setup, include=FALSE, message=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(tidytext)
library(data.table)
library(httr)
library(leaflet)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
library(htmlTable)


# INITALIZE CODE CHUNK OPTIONS
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = FALSE,
  cache = FALSE,
  tidy = TRUE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
  class.source = "code-r")
```


```{css, echo = FALSE}
.code-r { /* Code block */
  font-size: 15px;
}

.code-r-small { /* Code block */
  font-size: 10px;
}
```
<br>

# Intro
  
  This project analyzed the leading cause of death in United States changing pattern regarding time, region and cause from 1999 to 2017. The purpose of this study is to assess what cause of death we should pay more attention to and generate possible hypotheses/explanation for furthur analysis.
  The research questions and hypothesis for this project to examine and analyze are:
  
  * [*Hypothesis*] The death rate of all-cause death rate follows a decreasing trend in the whole United States and each state due to improved medical car.
  * [*Questions*] What are the death rate changing patterns regarding time, region, and causes? What should we learn from these patterns?

<br>

----

# Data

  The dataset used for this project presents **the age-adjusted death rates** for the 10 leading causes of death in the United States beginning from 1999 to 2017. Causes of death are classified by the [International Classification of Diseases](https://www-doh.state.nj.us/doh-shad/query/NCHS113.htm). Dateset is accessed from CDC (https://data.cdc.gov/NCHS/NCHS-Leading-Causes-of-Death-United-States/bi63-dtpu)
  
  Other dataset included in this analysis:
  
  * The United States regions dataset. It contains all 51 states, and groups them into 5 regions according to their geographic position on the continent: the Northeast, Southwest, West, Southeast, and Midwest. ([ref](https://www.nationalgeographic.org/maps/united-states-regions/) 
[data source](https://www.kaggle.com/omer2040/usa-states-to-region))
  * The US state [location dataset](https://www.kaggle.com/washimahmed/usa-latlong-for-state-abbreviations)
  
  EDA checklist was applied to clean and wrangle our data. Here is some basic information about this dataset. 

N=10868 overservation in total. 6 variables included:

1) Year
2) 113 Cause Name: International classification of Disease
3) Cause Name: a briefly version of 113 Cause Name, only have 10 leading cause of death here in this dataset
4) State: 52 Unique values include 51 states and the record of the US
5) Deaths: Death number
6) Age-adjusted Death Rate(AADR): Age-adjusted death rates (per 100,000 population) Adjusted based on the 2000 U.S. standard population. Populations used for computing death rates after 2010 are postcensal estimates based on the 2010 census, estimated as of July 1, 2010.

```{r data loading, cache=TRUE}
if (!file.exists("bi63-dtpu.csv"))
  download.file(
    url = "https://data.cdc.gov/api/views/bi63-dtpu/rows.csv",
    destfile = "bi63-dtpu.csv",
    method   = "libcurl",
    timeout  = 60
    )
data <- data.table::fread("bi63-dtpu.csv")

region <- data.table::fread("https://raw.githubusercontent.com/NJY530/pm566-project/main/Midterm/USAregion2.csv")

location <- data.table::fread("https://raw.githubusercontent.com/NJY530/pm566-project/main/Midterm/USAlocation.csv")
```


```{r checking NA, eval=FALSE}
d1 = summary(is.na(data))
d2 = summary(data)
knitr::kable(rbind(d1,d2))
```


```{r merge data}
d_region<-merge(
  x=data,
  y=region,
  by.x= "State",
  by.y = "State",
  all.x = TRUE,
  all.y = FALSE
)

alldata <- merge(
  x=d_region,
  y=location,
  by.x = "State",
  by.y = "City",
  all.x=TRUE,
  all.y=FALSE
)

```


<br>

----


# Data exploration and Preliminary Results

In this section, we will explore the death rate changing pattern from three perspectives: time, region and leading cause.

<br>

## Time

Use `Plotly` to draw line chart (year vs death rate) and grouped by states.

```{r all cause dataset}
allcause<- alldata %>%
  filter(alldata$`Cause Name` == "All causes")
```

```{r line chart}
linechart <- ggplot(allcause) +
  geom_line(mapping = aes(x=allcause$`Year`, y = allcause$`Age-adjusted Death`, color=State))+
  labs(title = "Death rate change in each state and the whole US from 1999 to 2017",x="year", y ="death rate per 100,000", size = 10)
ggplotly(linechart)
```

This graph indicates all 51 states and the whole United states' *age-adjusted death rate* change during 18 years. Hover indicates specific death rate data and state information. According to this graph, all 51 states follow a general downward trend which meets our hypothesis that as time changes, the level of medical care has improved, and the mortality rate has shown a downward trend.

<br>

## Region {.tabset}


```{r 1999 data}
#Extract data for each state in 1999
data1999<- alldata %>%
  group_by(State) %>%
  filter(Year ==1999, `Cause Name` == "All causes")%>%
  select(State,`Age-adjusted Death Rate`, Deaths, `State Code`)

#Create hover text
data1999$hover <- with(data1999, paste(State, '<br>', "Age-adjusted Death Rate: ", `Age-adjusted Death Rate`, '<br>', "All-cause Death number: ", Deaths))

#Set up mapping details
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

fig <- plot_geo(data1999, locationmode = 'USA-states') %>% 
  add_trace(
    z = ~`Age-adjusted Death Rate`, text = ~hover, locations = ~`State Code`,
    color = ~`Age-adjusted Death Rate`, colors = 'Reds'
  )

fig <- fig %>% colorbar(title = "Age-adjusted Death Rate in 1999", limits = c(550,1100))
fig <- fig %>% layout(
    title = 'Age-adjusted Death Rate in 1999',
    geo = set_map_details
  )

fig_1999 <- fig

```


```{r map2017}
#Extract data for each state in 2017
data2017<- alldata %>%
  group_by(State) %>%
  filter(Year ==2017, `Cause Name` == "All causes")%>%
  select(State,`Age-adjusted Death Rate`, Deaths, `State Code`)

#Create hover text
data2017$hover <- with(data2017, paste(State, '<br>', "Age-adjusted Death Rate: ", `Age-adjusted Death Rate`, '<br>', "All-cause Death number: ", Deaths))

#Set up mapping details
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

fig <- plot_geo(data2017, locationmode = 'USA-states') %>% 
  add_trace(
    z = ~`Age-adjusted Death Rate`, text = ~hover, locations = ~`State Code`,
    color = ~`Age-adjusted Death Rate`, colors = 'Reds'
  )

fig <- fig %>% colorbar(title = "Age-adjusted Death Rate in 2017", limits = c(550,1100))
fig <- fig %>% layout(
    title = 'Age-adjusted Death Rate in 2017',
    geo = set_map_details
  )

fig_2017 <- fig

```

### Age-adjusted Death Rate Map in 1999

```{r echo=FALSE}
fig_1999
```

### Age-adjusted Death Rate Map in 2017

```{r echo=FALSE}
fig_2017
```

## {-}

These two maps illustrate the *Age-adjusted Death Rate* distribution in 1999 and 2017 across the US. Color represents the death rate. The higher the death rate, the darker the red is. Hover includes AADR and death number.

We can see from these two graphs that compared with 1999, the death rate in 2017 has dropped significantly in every state. It's worth noticing that the entire southeast region, including Missouri, Louisiana, Oklahoma, Mississippi, Kentucky, West Virginia, Tennessee, etc., has the highest mortality rate in both 1999 and 2017. Possible explanation analysis see the [supplement section]().

<br>

## Causes

- Create a scatter plot representing `Year` vs. `Deaths` in the US.

```{r scatter plot}
#Extract data for the whole US from 1999 to 2017
US_cause <- data %>%
  select(State, Year, `Cause Name`,`Deaths`,`Age-adjusted Death Rate`) %>%
  filter(State == "United States", `Cause Name` != "All causes")
  

US_cause %>% 
  plot_ly(x = ~Year, y = ~`Age-adjusted Death Rate`, 
          type = 'scatter', mode = 'markers', color = ~`Cause Name`,
          size = ~`Deaths`, sizes = c(5, 30), marker = list(sizemode='diameter', opacity=0.5),
          hoverinfo = 'text',
          text = ~paste( paste(`Cause Name`, ":", sep=""), paste(" Death Number ", Deaths, sep="") , paste(" Age-adjusted Death Rate: ", `Age-adjusted Death Rate`, sep=""), sep = "<br>")) %>%
  add_lines(y=~`Age-adjusted Death Rate`)%>%
  layout(title = "Yeas vs. Age-adjusted Death rate in respct of leading cause ",
                  yaxis = list(title = "Age-adjusted Death Rate"), xaxis = list(title = "Year"),
         hovermode = "compare")

```

This scttor plot indicates *age-adjusted death rate* changes by *year*. Bubble size is based on *death number*. Each different cause of death presented as different color. Hover includes the disease that causes death, death rate, and death number.

We can see that in this graph, except for heart disease and cancer, the death rates of the other eight causes of death in 18 years are all lower than 60, fluctuating within a specific range. Although they have been declining, the death rates from cancer and heart disease are still three to four times that of other causes of death. This means that heart disease and cancer are still the predominant leading causes of death in the United States.

It is worth noting that when we carefully observe these data. For most causes of death, the number of deaths is monotonously increasing rather than monotonously decreasing, such as cancer, Alzheimer's disease, etc. The reason for this may be population growth and population ageing.

<br>

----

# Conclusion

Now we could draw a preliminary conclusion to our question:

It's apparent that all death rates went down in the past 18 years in all 51 states. Compared to 1999, the death rate decreased in 2017. The regional distribution of death rates did not show significant differences. However, it is worth noting that the mortality rate in the southeast region is higher than that in other regions, and the authority should pay attention to the causes of these abnormal mortality rates.

From the disease cause aspect, in 1999, the fatal death cause was heart disease, while in 2017, cancer caught up a little. This might indicate that nowadays, cancer should pay more attention to. In response to the decline in the death rate but the increase in the number of deaths, I hypothesize that such a trend is due to the increase in medical standards, leading to a significant decline in mortality, which is consistent with our previous assumptions. The increase in the number of deaths may represent an increase in the number of people and the current trend of population aging.

<br>

----

# Download the report

The PDF version could be downloaded [here]()

<br>

----

# Reference


<br>